Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-10-02 src/tex/ms.tex
add figure

Aligned-spin, frequency-domain waveform models (such as IMRPhenomD) can be written as a
combination of amplitude and phase functions ($A$ and $\phi$ respectively):
\begin{align}\label{eq:}
	h(f,\theta,\Lambda) = A(f,\theta,\Lambda)e^{-i\phi(f,\theta,\Lambda^i)}\,,
\end{align}
where $f$ is the frequency, $\theta$ are the intrinsic parameters of the binary, and $\Lambda^i$ \rr{($i=1,2,\cdots,19$) is a set of 19 additional
parameters} which will be discussed below. 
The phase and amplitude functions are then split into three sections which represent the
inspiral, intermediate, and merger-ringdown (MR) parts of the waveform. 
During inspiral, $A$ and $\phi$ are known analytically from post-Newtonian (PN) theory;
IMRPhenomD uses the TaylorF2 model \citep{Buonanno:2009zt, Arun:2004hn} which is known up to 3.5PN order.
To model the intermediate and MR regions, IMRPhenomD (and all waveforms in the IMRPhenom family)
uses a series of parameterizations\footnote{
	The parameterizations for both the amplitude and phase functions can be found in \citep{khan2016frequency}.
} 
which depend purely on $\Lambda^i$ and can be calibrated to numerical relativity (NR) simulations.
The three sections are then \textit{stitched} together using step functions.
Importantly, the parameterizations are chosen such that they can be made $\mathcal{C}^1$ continuous at the
boundary between each section.

@@

With the purpose of improving downstream tasks such as parameter estimation in
mind, the more relevant metric of improvement is the distribution of improvement
in mismatch over the entire validation dataset. Figure~\ref{fig:q148} shows the
distribution of log-mismatches for the validation waveforms before and after the
optimization procedure. Here we show results using a constant PSD in our loss
function. One can see the distribution of the optimized waveform is skewed toward lower mismatches,
with the peak of the distribution being shifted by approximately an order of
magnitude, and the median mismatch being reduced by 50\% (see vertical dotted lines). When using
$\mathcal{L}_{\mathrm{norm}}$, we observe a less pronounced improvement with a 22.9\%
decrease in the median.

@@

\kl{We note that in comparison to the original IMRPhenomD calibration, we restricted the 
parameter space to $q\leq8$ and directly used NR waveform instead of hybridized SEOB-NR 
waveforms, which clearly put the original IMRPhemonD result in a 
disadvantage. We have tried to train the model using a set of training waveforms of the same 
intrinsic parameters, but with BAM waveforms replaced with SEOBNR waveforms. The result 
shows a marginal improvement, which corresponds to the stitching problem discussed before.
However, it is known that IMRPhenomD has a large inaccuracy in the large mass-ratio regime, 
and is rarely used in this region of parameter space. Hence, we focus on showing AD with 
gradient descent, as a general and robust method, can be used to further optimize models 
independent of the training waveforms used.}

>>>>

Aligned-spin, frequency-domain waveform models (such as IMRPhenomD) can be written as a
combination of amplitude and phase functions ($A$ and $\phi$ respectively):
\begin{align}\label{eq:}
	h(f,\theta,\Lambda) = A(f,\theta,\Lambda)e^{-i\phi(f,\theta,\Lambda^i)}\,,
\end{align}
where $f$ is the frequency, $\theta$ are the intrinsic parameters of the binary, and $\Lambda^i$ \rr{($i=1,2,\cdots,19$) is a set of 19 additional
parameters} which will be discussed below. 
The phase and amplitude functions are then split into three sections which represent the
inspiral, intermediate, and merger-ringdown (MR) parts of the waveform. 
During inspiral, $A$ and $\phi$ are known analytically from post-Newtonian (PN) theory;
IMRPhenomD uses the TaylorF2 model \citep{Buonanno:2009zt, Arun:2004hn} which is known up 
to 3.5PN order \kl{augmented with the next four higher order PN terms}.
To model the intermediate and MR regions, IMRPhenomD (and all waveforms in the IMRPhenom family)
uses a series of parameterizations\footnote{
	The parameterizations for both the amplitude and phase functions can be found in \citep{khan2016frequency}.
} 
which depend purely on $\Lambda^i$ and can be calibrated to numerical relativity (NR) simulations.
The three sections are then \textit{stitched} together using step functions.
Importantly, the parameterizations are chosen such that they can be made $\mathcal{C}^1$ continuous at the
boundary between each section.

@@

With the purpose of improving downstream tasks such as parameter estimation in
mind, the more relevant metric of improvement is the distribution of improvement
in mismatch over the entire validation dataset. Figure~\ref{fig:q148} shows the
distribution of log-mismatches for the validation waveforms before and after the
optimization procedure. Here we show results using a constant PSD in our loss
function. One can see the distribution of the optimized waveform is skewed toward lower mismatches,
with the peak of the distribution being shifted by approximately an order of
magnitude. The median mismatch is reduced by 50\% (see vertical dotted lines) \kl{while the tail hardly changes}. When using
$\mathcal{L}_{\mathrm{norm}}$, we observe a less pronounced improvement with a 22.9\%
decrease in the median.

@@

\kl{We note that in comparison to the original IMRPhenomD calibration, we restricted the 
parameter space to $q\leq8$ and directly used NR waveform instead of hybridized SEOB-NR 
waveforms in training and validation purposes. These choices put the original IMRPhemonD 
result in a disadvantage, as both the parameter space and frequency range are restricted. 
We tested our result by training the model using a set of training waveforms of the same 
intrinsic parameters, but with BAM waveforms replaced with SEOBNR waveforms. We can see 
the result in Fig.~\ref{fig:q14818}, where the mismatch median changes by around $8\%$, 
This improvement mainly corresponds to the stitching problem discussed before. 
Nevertheless, it is known that IMRPhenomD has a large inaccuracy in the large mass-ratio 
regime, and is rarely used in this region of parameter space. Hence, we focus on showing 
AD with gradient descent, as a general and robust method, can be used to further optimize 
models independent of the training waveforms used.}

\begin{figure}[t]
	\script{q14818.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/q14818.pdf}
	\caption{Distributions of mismatches before and after optimization using the 
		$\mathcal{L}_{\mathrm{ave}}$ loss function. Mismatches are calculated using the training waveforms listed in \cite{khan2016frequency}, but BAM waveforms are replaced by SEOBNRv4 waveforms of the same intrinsic parameters. 
		Mismatches are calculated with a constant PSD. The dotted lines represent the median of the distributions, which decreased by $8.2\%$ during optimization.}
	\label{fig:q14818}
\end{figure}
